Equations & R Code

### 2017-04-03 manuscript/coexistence-stability_ms.Rmd
minor deletions to get word count under5000

@@

\noindent \textbf{Number of Words in Abstract}: xxx

\noindent \textbf{Number of Words in Main Text}: xxxx

\noindent \textbf{Number of References}: xx

@@

\begin{abstract}
Theory relating species richness to ecosystem variability typically ignores the potential for environmental variability to promote species coexistence.
Failure to account for fluctuation-dependent coexistence mechanisms may explain observed deviations from the expected negative diversity--\new{ecosystem variability} relationship, and limits our ability to predict the consequences of future increases in environmental variability.
We use a consumer-resource model to explore how coexistence via the temporal storage effect and relative nonlinearity affects ecosystem variability.
We show that a positive, rather than negative, diversity--\new{ecosystem variability} relationship is possible when ecosystem function is sampled across a natural gradient in environmental variability and diversity.
We also show how fluctuation-dependent coexistence can buffer ecosystem functioning against increasing environmental variability by promoting species richness and portfolio effects.
Our work provides a general explanation for variation in observed diversity--\new{ecosystem variability} relationships and highlights the importance of conserving regional species pools to help buffer ecosystems against predicted increases in environmental variability.
\vspace{2em}
\end{abstract}

@@

To reconcile these two perspectives, we extend theory on the relationship between species richness and ecosystem variability to cases in which species coexistence explicitly depends on environmental fluctuations and species-specific responses to environmental conditions.
We focus on \new{communities where coexistence is maintained by either the temporal storage effect or relative nonlinearity, but not both,} using a general consumer-resource model.
We use the model to investigate two questions:

@@

\new{We limit our analysis to four-species communities because it is exceedingly difficult to get more than four species to coexist via relative nonlinearity without introducing some form of fluctuation-independent mechanism like resource partitioning} [@Yuan2015], \new{which would then cloud our inference on the effect of fluctuation-dependent coexistence.
For consistency, we also constrain our focus to four species communities under the storage effect, but our conclusions apply to more species-rich communities (see Supporting Information section SI.2).}

@@

### Numerical simulations
To explore how fluctuation-dependent coexistence can affect the diversity--ecosystem variability relationship, we simulated the model with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many species can coexist, akin to a community assembly experiment with a species pool of four species.
\new{To do this, we simulated communities with all species initially present across a gradient of annual resource variability for relative nonlinearity (50 evenly-spaced values of $\sigma_R$ in the range [0, 1.2]) or environmental cue variability for the storage effect (100 evenly-spaced values of $\sigma_E^2$ in the range [0, 3]).}
Second, we chose parameter values that allowed coexistence of all four species and then performed species removals \new{at a single level of environmental variability}, akin to a biodiversity--ecosystem function experiment.
The two simulation experiments correspond to (i) sampling ecosystem function across a natural gradient of species richness and (ii) sampling ecosystem function across diversity treatments within a site.
\new{We refer to the former as a "regional" relationship, and the latter as a "local" relationship.
But note that we do not attribute any particular area size to "region", it is simply any area over which a gradient of environmental variability can emerge.}

To understand how increasing environmental variability will impact ecosystem variability when coexistence is fluctuation-dependent, we simulated the model over a range of species pool sizes and environmental cue or resource variability.
For each size of species pool (1, 2, 3, or 4 species), we simulated the model at 15 evenly-spaced levels of environmental cue (range = [0.1, 2]) for the storage effect and 25 evenly-spaced levels of resource variability (range = [0, 1.2]) for relative nonlinearity.
We also explored the influence of asymmetries in species' competitive abilities and correlations in species' environmental responses within the storage effect model.
\new{We created competitive hierarchies by making the live-to-dormant biomass fractions ($\alpha$s) unequal among species.
A small difference among values of $\alpha$ were needed to create competitive hierarchies because we chose a relatively constrained gradient of environmental cue variance.
Larger differences among values of $\alpha$ expand the region of coexistence farther along a gradient environmental cue variance.
These parameter value choices do not affect our conclusions, but would change the quantitative results.}

@@

\new{For parameters that we did not vary, we chose values that would allow coexistence of all four species at some point along the environmental variability gradients we simulated.
Our focus is specifically on communities where fluctuation-dependent coexistence is operating, and making parameters increasingly asymmetric among species typically reduced coexistence strength or made coexistence impossible (Supporting Information section SI.3).
Changes in the absolute values of parameters also altered the strength of coexistence, but in no case did altering parameter values change the qualitative results and conclusions we present below (see Supporting Information section SI.3).}
Parameter values for specific results are given in figure captions.

@@

In communities where species coexist via relative nonlinearity, the extent to which species additions buffer ecosystem stability against increases in environmental variability depends on the species traits of immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are less intrinsically stable than the resident species, ecosystem variability increases at a relatively constant rate even as species are added (Fig. 4A; Fig. SI-11).
If more stable species colonize, species additions buffer the ecosystem from increasing environmental variability (Fig. 4B).
\new{The "stair-steps" in Fig. 4B emerge because a different species is making up the one-species community at different levels of resource variability, increasing ecosystem \emph{CV} each time as a more unstable species composes the single species community.}

\new{We tested the generality of our results under different parameters by conducting a targeted sensitivity analysis focused on parameter values and asymmetries that most affect species coexistence (Supporting Information section SI.3).
Making parameters asymmetric creates competitive hierachies and makes coexistence more difficult.
For the storage effect, this is evident in Fig. 3 where we made live-to-dormant transition rates ($\alpha$s) asymmetric, and the same result occurred when we made dormant mortality rates ($\eta$s) asymmetric (Fig. SI-2).
Furthermore, increasing the absolute value of dormant mortality also made coexistence more difficult, but results were qualitatively similar to those above (Fig. SI-3).
For relative nonlinearity, altering the absolute value of the mean resource pulse shifts the region of coexistence across a gradient of resource variability (Fig. SI-4).
But, again, qualitatively similar results emerge when species \emph{can} coexist.
In general, altering any parameter in isolation will make coexistence easier or harder at any given level of environmental variability.
Thus, our results are robust in the sense that the qualitative patterns we present will always emerge for parameter values that permit species coexistence over some range of envrionmental variability.
In other words, our results are only sensitive to whether or not fluctuation-dependent coexistence is operating.}

>>>>

\noindent \textbf{Data Accessibility}: There is no data associated with this manuscript. All R code necessary to reproduce our results will be archived on Figshare and released on GitHub (https://github.com/atredennick/Coexistence-Stability).

@@

\noindent \textbf{Number of Words in Abstract}: 150

\noindent \textbf{Number of Words in Main Text}: 4,978 (including in-text references)

\noindent \textbf{Number of References}: 48

@@

\begin{abstract}
Theory relating species richness to ecosystem variability typically ignores the potential for environmental variability to promote species coexistence.
Failure to account for fluctuation-dependent coexistence may explain deviations from the expected negative diversity--\new{ecosystem variability} relationship, and limits our ability to predict the consequences of increases in environmental variability.
We use a consumer-resource model to explore how coexistence via the temporal storage effect and relative nonlinearity affects ecosystem variability.
We show that a positive, rather than negative, diversity--\new{ecosystem variability} relationship is possible when ecosystem function is sampled across a natural gradient in environmental variability and diversity.
We also show how fluctuation-dependent coexistence can buffer ecosystem functioning against increasing environmental variability by promoting species richness and portfolio effects.
Our work provides a general explanation for variation in observed diversity--\new{ecosystem variability} relationships and highlights the importance of conserving regional species pools to help buffer ecosystems against predicted increases in environmental variability.
\vspace{2em}
\end{abstract}

@@

To reconcile these two perspectives, we extend theory on the relationship between species richness and ecosystem variability to cases in which species coexistence explicitly depends on environmental fluctuations and species-specific responses to environmental conditions.
We focus on \new{communities where coexistence is maintained by either the temporal storage effect or relative nonlinearity} using a general consumer-resource model.
We use the model to investigate two questions:

@@

\new{We limit our analysis to four-species communities because it is exceedingly difficult to get more than four species to coexist via relative nonlinearity without introducing another coexistence mechanism} [@Yuan2015].
For consistency, we also constrain our focus to four species communities under the storage effect, but our conclusions apply to more species-rich communities (see Supporting Information section SI.2).}

@@

### Numerical simulations
To explore how fluctuation-dependent coexistence can affect the diversity--ecosystem variability relationship, we simulated the model with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many species can coexist, akin to a community assembly experiment with a species pool of four species.
\new{We simulated communities with all species initially present across a gradient of annual resource variability for relative nonlinearity (50 evenly-spaced values of $\sigma_R$ in the range [0, 1.2]) or environmental cue variability for the storage effect (100 evenly-spaced values of $\sigma_E^2$ in the range [0, 3]).}
Second, we chose parameter values that allowed coexistence of all four species and then performed species removals \new{at a single level of environmental variability}, akin to a biodiversity--ecosystem function experiment.
The two simulation experiments correspond to (i) sampling ecosystem function across a natural gradient of species richness and (ii) sampling ecosystem function across diversity treatments within a site.
\new{We refer to the former as a "regional" relationship, and the latter as a "local" relationship.
But we do not attribute any particular area size to "region", it is simply any area over which a gradient of environmental variability exists.}

To understand how increasing environmental variability will impact ecosystem variability when coexistence is fluctuation-dependent, we simulated the model over a range of species pool sizes and environmental cue or resource variability.
For each size of species pool (1, 2, 3, or 4 species), we simulated the model at 15 evenly-spaced levels of environmental cue (range = [0.1, 2]) for the storage effect and 25 evenly-spaced levels of resource variability (range = [0, 1.2]) for relative nonlinearity.
We also explored the influence of asymmetries in species' competitive abilities and correlations in species' environmental responses within the storage effect model.
\new{We created competitive hierarchies by making the live-to-dormant biomass fractions ($\alpha$s) unequal among species.
Small differences among values of $\alpha$ were needed to create competitive hierarchies because we chose a relatively constrained gradient of environmental cue variance.
Larger differences among values of $\alpha$ expand the region of coexistence farther along a gradient environmental cue variance.}

@@

\new{For parameters that we did not vary, we chose values that would allow coexistence of all four species at some point along the environmental variability gradients we simulated.
Our focus is specifically on communities where fluctuation-dependent coexistence is operating, and making parameters increasingly asymmetric among species typically reduced coexistence strength or made coexistence impossible (Supporting Information section SI.3).
Changes in the absolute values of parameters also altered the strength of coexistence, but in no case did altering parameter values change our qualitative results and conclusions (Supporting Information section SI.3).}
Parameter values for specific results are given in figure captions.

@@

In communities where species coexist via relative nonlinearity, the extent to which species additions buffer ecosystem stability against increases in environmental variability depends on the species traits of immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are less intrinsically stable than the resident species, ecosystem variability increases at a relatively constant rate even as species are added (Fig. 4A; Fig. SI-11).
If more stable species colonize, species additions buffer the ecosystem from increasing environmental variability (Fig. 4B).
<!---
\new{The "stair-steps" in Fig. 4B emerge because a different species is making up the one-species community at different levels of resource variability, increasing ecosystem \emph{CV} each time as a more unstable species composes the single species community.}
--->

\new{We tested the generality of our results under different parameters by conducting a targeted sensitivity analysis focused on parameter values and asymmetries that most affect species coexistence (Supporting Information section SI.3).
In general, altering any parameter in isolation will make coexistence easier or harder at any given level of environmental variability.
Our results are only sensitive to whether or not fluctuation-dependent coexistence is operating.}

<!---
Making parameters asymmetric creates competitive hierachies and makes coexistence more difficult.
For the storage effect, this is evident in Fig. 3 where we made live-to-dormant transition rates ($\alpha$s) asymmetric, and the same result occurred when we made dormant mortality rates ($\eta$s) asymmetric (Fig. SI-2).
Furthermore, increasing the absolute value of dormant mortality also made coexistence more difficult, but results were qualitatively similar to those above (Fig. SI-3).
For relative nonlinearity, altering the absolute value of the mean resource pulse shifts the region of coexistence across a gradient of resource variability (Fig. SI-4).
But, again, qualitatively similar results emerge when species \emph{can} coexist.

Thus, our results are robust in the sense that the qualitative patterns we present will always emerge for parameter values that permit species coexistence over some range of envrionmental variability.
In other words, }
--->

### 2017-04-03 manuscript/coexistence-stability_ms.tex
moved figures to end; submitted version
minor deletions to get word count under5000

\new{We limit our analysis to four-species communities because it is exceedingly difficult to get more than four species to coexist via relative nonlinearity without introducing some form of fluctuation-independent mechanism like resource partitioning}
(Yuan \& Chesson 2015),
\new{which would then cloud our inference on the effect of fluctuation-dependent coexistence.
For consistency, we also constrain our focus to four species communities under the storage effect, but our conclusions apply to more species-rich communities (see Supporting Information section SI.2).}

@@

To explore how fluctuation-dependent coexistence can affect the
diversity--ecosystem variability relationship, we simulated the model
with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many
species can coexist, akin to a community assembly experiment with a
species pool of four species.
\new{To do this, we simulated communities with all species initially present across a gradient of annual resource variability for relative nonlinearity (50 evenly-spaced values of $\sigma_R$ in the range [0, 1.2]) or environmental cue variability for the storage effect (100 evenly-spaced values of $\sigma_E^2$ in the range [0, 3]).}
Second, we chose parameter values that allowed coexistence of all four
species and then performed species removals
\new{at a single level of environmental variability}, akin to a
biodiversity--ecosystem function experiment. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.
\new{We refer to the former as a "regional" relationship, and the latter as a "local" relationship.
But note that we do not attribute any particular area size to "region", it is simply any area over which a gradient of environmental variability can emerge.}

To understand how increasing environmental variability will impact
ecosystem variability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = {[}0.1, 2{]}) for the storage effect and 25
evenly-spaced levels of resource variability (range = {[}0, 1.2{]}) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model.
\new{We created competitive hierarchies by making the live-to-dormant biomass fractions ($\alpha$s) unequal among species.
A small difference among values of $\alpha$ were needed to create competitive hierarchies because we chose a relatively constrained gradient of environmental cue variance.
Larger differences among values of $\alpha$ expand the region of coexistence farther along a gradient environmental cue variance.
These parameter value choices do not affect our conclusions, but would change the quantitative results.}

@@

In communities where species coexist via relative nonlinearity, the
extent to which species additions buffer ecosystem stability against
increases in environmental variability depends on the species traits of
immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are
less intrinsically stable than the resident species, ecosystem
variability increases at a relatively constant rate even as species are
added (Fig. 4A; Fig. SI-11). If more stable species colonize, species
additions buffer the ecosystem from increasing environmental variability
(Fig. 4B).
\new{The "stair-steps" in Fig. 4B emerge because a different species is making up the one-species community at different levels of resource variability, increasing ecosystem \emph{CV} each time as a more unstable species composes the single species community.}

\new{We tested the generality of our results under different parameters by conducting a targeted sensitivity analysis focused on parameter values and asymmetries that most affect species coexistence (Supporting Information section SI.3).
Making parameters asymmetric creates competitive hierachies and makes coexistence more difficult.
For the storage effect, this is evident in Fig. 3 where we made live-to-dormant transition rates ($\alpha$s) asymmetric, and the same result occurred when we made dormant mortality rates ($\eta$s) asymmetric (Fig. SI-2).
Furthermore, increasing the absolute value of dormant mortality also made coexistence more difficult, but results were qualitatively similar to those above (Fig. SI-3).
For relative nonlinearity, altering the absolute value of the mean resource pulse shifts the region of coexistence across a gradient of resource variability (Fig. SI-4).
But, again, qualitatively similar results emerge when species \emph{can} coexist.
In general, altering any parameter in isolation will make coexistence easier or harder at any given level of environmental variability.
Thus, our results are robust in the sense that the qualitative patterns we present will always emerge for parameter values that permit species coexistence over some range of envrionmental variability.
In other words, our results are only sensitive to whether or not fluctuation-dependent coexistence is operating.}

>>>>

\new{We limit our analysis to four-species communities because it is exceedingly difficult to get more than four species to coexist via relative nonlinearity without introducing another coexistence mechanism}
(Yuan \& Chesson 2015). For consistency, we also constrain our focus to
four species communities under the storage effect, but our conclusions
apply to more species-rich communities (see Supporting Information
section SI.2).\}

@@

To explore how fluctuation-dependent coexistence can affect the
diversity--ecosystem variability relationship, we simulated the model
with four species under two scenarios for each coexistence mechanism.
First, we allowed the variance of the environment to determine how many
species can coexist, akin to a community assembly experiment with a
species pool of four species.
\new{We simulated communities with all species initially present across a gradient of annual resource variability for relative nonlinearity (50 evenly-spaced values of $\sigma_R$ in the range [0, 1.2]) or environmental cue variability for the storage effect (100 evenly-spaced values of $\sigma_E^2$ in the range [0, 3]).}
Second, we chose parameter values that allowed coexistence of all four
species and then performed species removals
\new{at a single level of environmental variability}, akin to a
biodiversity--ecosystem function experiment. The two simulation
experiments correspond to (i) sampling ecosystem function across a
natural gradient of species richness and (ii) sampling ecosystem
function across diversity treatments within a site.
\new{We refer to the former as a "regional" relationship, and the latter as a "local" relationship.
But we do not attribute any particular area size to "region", it is simply any area over which a gradient of environmental variability exists.}

To understand how increasing environmental variability will impact
ecosystem variability when coexistence is fluctuation-dependent, we
simulated the model over a range of species pool sizes and environmental
cue or resource variability. For each size of species pool (1, 2, 3, or
4 species), we simulated the model at 15 evenly-spaced levels of
environmental cue (range = {[}0.1, 2{]}) for the storage effect and 25
evenly-spaced levels of resource variability (range = {[}0, 1.2{]}) for
relative nonlinearity. We also explored the influence of asymmetries in
species' competitive abilities and correlations in species'
environmental responses within the storage effect model.
\new{We created competitive hierarchies by making the live-to-dormant biomass fractions ($\alpha$s) unequal among species.
Small differences among values of $\alpha$ were needed to create competitive hierarchies because we chose a relatively constrained gradient of environmental cue variance.
Larger differences among values of $\alpha$ expand the region of coexistence farther along a gradient environmental cue variance.}

@@

In communities where species coexist via relative nonlinearity, the
extent to which species additions buffer ecosystem stability against
increases in environmental variability depends on the species traits of
immigrating species and the order in which they enter the community.
When additional species, which immigrate from the regional pool, are
less intrinsically stable than the resident species, ecosystem
variability increases at a relatively constant rate even as species are
added (Fig. 4A; Fig. SI-11). If more stable species colonize, species
additions buffer the ecosystem from increasing environmental variability
(Fig. 4B).

\new{We tested the generality of our results under different parameters by conducting a targeted sensitivity analysis focused on parameter values and asymmetries that most affect species coexistence (Supporting Information section SI.3).
In general, altering any parameter in isolation will make coexistence easier or harder at any given level of environmental variability.
Our results are only sensitive to whether or not fluctuation-dependent coexistence is operating.}
